### 2023-08-10 chapters/conclusion.tex
thesis conclusio

Our work on fairness in machine learning introduces two novel probabilistic fairness verifiers, {\justicia} and {\fvgm}, which exhibit superior performance in accuracy and scalability compared to state-of-the-art verifiers. {\justicia} is a stochastic-SAT-based verifier that enables scalable verification of fairness for compound sensitive groups of Boolean classifiers, such as decision trees, which was previously infeasible with existing methods. On the other hand, {\fvgm} takes correlated features into account and is capable of verifying the fairness of linear classifiers with higher scalability and accuracy than previous verifiers. Additionally, we discuss a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features. Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local explainability-based approach.

In future, our research on fair and interpretable machine learning will focus on several areas, which we elaborate in the following.

\begin{itemize}
	\item 
	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains. In this thesis, we have designed interpretable classifiers that are explainable by design. A natural application of interpretable models is to apply them to explain the working of black-box classifiers. For example, rule-based classifiers can be leveraged to provide explanations in the form of classification rules, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local}. As such, we can use interpretable classifiers {\imli} and {\crr} to generate post-hoc explanations for black-box classifiers.  In future, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.

	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, even beyond machine learning. This can help us scale up solutions without losing much accuracy. For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets, where the classifier can be routinely updated with new samples. Furthermore, we aim to try different objective functions in the incremental solving, apart from the hamming distance-based heuristic objective, which  works well in the incremental learning of CNF classification rules. One idea is to learn different rule-based classifiers, possibly in parallel, on different mini-batches or on different subsets of features, and then construct a rule merging procedure to combine different classification rules into one.

@@

		Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability. We plan to explore more directions in the future.

	\end{enumerate}

\end{itemize}

To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond. To achieve this goal, we aim to integrate formal methods and automated reasoning with statistics and algorithm design.

\begin{comment}
	\begin{itemize}
		\item Fairness repair
		\item Incremental solving
		\item Fairness and interpretability as a service to more complex models.
	\end{itemize}
\end{comment}

>>>>

Our work on fairness in machine learning introduces two novel probabilistic fairness verifiers, {\justicia} and {\fvgm}, which exhibit superior performance in accuracy and scalability compared to state-of-the-art verifiers. {\justicia} is a stochastic-SAT-based verifier that enables scalable verification of fairness for compound sensitive groups of Boolean classifiers, such as decision trees, which was previously infeasible with existing methods. On the other hand, {\fvgm} takes correlated features into account and is capable of verifying the fairness of linear classifiers with higher scalability and accuracy than previous verifiers. Additionally, we discuss a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features. Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local interpretability-based approach.

Our future research is dedicated to developing practical and scalable algorithms for trustworthy machine learning. Machine learning and artificial intelligence have been compared to the new electricity, with the potential to transform various aspects of human life, evident from the overwhelming response to generative AI. Ensuring fairness and interpretability in deployed machine learning is now more necessary than ever. To accomplish this, we aim to work in a collaborative environment, gaining insights into real-world challenges and leveraging advances in the field, alongside formal methods, to make significant progress. We have identified key research themes that will guide our work towards this vision.

\paragraph{Fairness and Interpretability As a Service.} 
The goal of modern machine learning extends beyond learning patterns from large-scale historical data to ensuring responsible decision-making through careful regulation to establish trustworthiness. For instance, in a job application scenario, a machine learning algorithm must be fair across different demographic groups, resilient to non-actionable changes in candidate profiles, and interpretable to allow candidates to understand the decision-making process. The long-term research goal of this thesis is to offer fairness and interpretability as a service with machine learning-based decision-making.

\begin{itemize}

	\item \textbf{Interpretability with Guarantees.} Our research in interpretable machine learning spans \textbf{two-fold directions}.  (i) \textit{Interpretability by design:} There is a growing interest for interpretable machine learning in safety-critical domains~\cite{rudin2019stop}. Building upon our interpretable rule-based classifier $ \mathsf{IMLI} $, we aim to enhance learning algorithms for interpretable models in large-scale datasets across supervised, semi-supervised, and unsupervised settings. (ii) \textit{Post-hoc interpretability:} To explain black-box predictions, we focus on explanations with formal guarantees~\cite{ignatiev2019abduction,ribeiro2018anchors,guidotti2018local}. For example, an explanation model must be robust, learned in a privacy-preserving manner, and provide the confidence level of explanations to increase transparency and trust in the decision-making process.

@@

	\end{enumerate}
\end{itemize}

\paragraph{Verifiable Machine Learning with Formal Methods.} In safety-critical and high-stake domains, the verification of machine learning before deployment is crucial. To this end, formal methods provide a template to verify different checkable properties of machine learning. In particular, SAT, SMT and their variants allow to concentrate on synthesizing constraints from real-world use cases and delegate the solution finding to respective solvers, thanks to the dedicated community in formal methods. Building upon our ongoing research, which involves applying SSAT and MaxSAT to address fairness and interpretability challenges, we plan to explore additional formulations in formal methods, such as functional analysis, abstract interpretation, and solvers with expressive theory, to further enhance the verification of machine learning models.

\paragraph{Benchmarking Machine Learning: Improvement in Formal Methods.} Interdisciplinary research is an ongoing process that yields valuable contributions to different disciplines. Our work on fairness and interpretability in machine learning has had a positive impact on the improvement of solvers in formal methods. For instance, when tackling interpretable classification problems, we realized that MaxSAT alone was insufficient for large-scale rule-based classification, prompting the need for an incremental MaxSAT solver. As a result, we contribute to the MaxSAT evaluation competition in 2019 with our MaxSAT benchmarks of interpretable classification. Subsequently, the competition introduced an incremental MaxSAT solving track. This observation serves as strong motivation for us to continue contributing to the formal methods community by creating additional machine learning verification benchmarks.
